dispRity manual

### 2020-07-16 inst/gitbook/dispRity_manual.Rmd
book update

### So which version do I choose? {#version}

Note that when we (the `dispRity` package collaborators!) are developing some more major versions, including some more complex features for example, they will never be released on either the the versions described above.
Therefore you can rest assure that you will never use a badly bugged version of `dispRity`! Enjoy!

@@

<!-- ## Associated papers -->

@@

@@

```{r, eval=TRUE}
## Generating four time slices every 40 million years under a model of proximity evolution
chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)

@@

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$area` |
1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
1 | `mode.val` | The modal value | `dispRity` |
1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
2 | `radius` | The radius of each dimensions | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
2 | `variances` | The variance of each dimension | `dispRity` |

@@

\begin{equation}
    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
\end{equation}

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}

\begin{equation}
    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
\end{equation}

\begin{equation}
    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
\end{equation}

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}

\begin{equation}
    span.tree.length = \sum(\mathrm{branch\ length})
\end{equation}

Where *k* is the number of dimensions,
*n* the number of elements,
$\Gamma$ is the Gamma distribution,
$\lambda_i$ is the eigenvalue of each dimensions,
$\sigma^{2}$ is their variance and
$Centroid_{k}$ is their mean,
$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).

@@

## Calculating disparity as the sum and the product of these variances
summary(dispRity(dummy_space, metric = c(sum, variances)))
summary(dispRity(dummy_space, metric = c(prod, variances)))

@@

#### Centroids and ancestral distance metrics

@@

```{r}
## The distance between each element and the origin of the ordinated space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))

## Disparity as the distance between each element and a specific point in space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
```

@@

@@

@@

## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Plotting the results
plot(test_unif, main = "Uniform (0,1)")
plot(test_norm1, main = "Normal (0,1)")
plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
plot(test_norm3, main = "Normal (variable + correlation)")

@@

@@

## The `dispRity` object content
The functions above are utilities to easily and safely access different elements in the `dispRity` object.
Alternatively, of course, each elements can be accessed manually.
Here is an explanation on how it works.
The `dispRity` object is a `list` of two to four elements, each of which are detailed below:

 * `$matrix`: an object of class `matrix`, the full multidimensional space.
 * `$call`: an object of class `list` containing information on the `dispRity` object content.
 * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
 * `$disparity`: an object of class `list` containing the disparity values.

The `dispRity` object is loosely based on `C` structure objects.
In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.

>>>>

@@

### Why is it important to cite us?

Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?

All the people involve in making the `dispRity` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
I created the package with this idea in mind and I am still sticking to it.
However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dispRity` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!

Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!

Thank you!
<!-- ## Associated papers -->

@@

### Time-slicing {#time-slicing}

@@

More details about the differences between these methods can be found in @time-slice.

```{r, eval=TRUE}
## Generating four time slices every 40 million years
## under a model of proximity evolution
chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
               method = "continuous", model = "proximity",
               time = c(120, 80, 40, 0),
               FADLAD = BeckLee_ages)

@@

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
2 | `angles` | The angle of main variation of each dimensions | `dispRity` |
2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$area` |
1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
2 | `deviations` | The minimal distance between each element and a hyperplane | `dispRity` |
1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
1 | `func.div` | The functional divergence (the ratio of deviation from the centroid) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
1 | `func.eve` | The functional evenness (the minimal spanning tree distances evenness) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
1 | `mode.val` | The modal value | `dispRity` |
1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
2 | `neighbours` | The distance to specific neighbours (e.g. the nearest neighbours - by default) | `dispRity` |
2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
2 | `radius` | The radius of each dimensions | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `span.tree.length` | The minimal spanning tree length | [`ape`](https://cran.r-project.org/web/packages/vegan/index.html) (but see also [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree`) |
2 | `variances` | The variance of each dimension | `dispRity` |

@@

\begin{equation}
    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({d}_{n}-Ancestor_{n})^2}}
\end{equation}

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{d})^2}}
\end{equation}

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{d}|max(d_i) - min(k_i)|}
\end{equation}

\begin{equation}
    deviations = \frac{|Ax + By + ... + Nm + Intercept|}{\sqrt{A^2 + B^2 + ... + N^2}}
\end{equation}

\begin{equation}
    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Reference_{d})^2}}}{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{k})^2}}}
\end{equation}

\begin{equation}
    ellipse.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} (\lambda_{i}^{0.5})
\end{equation}

\begin{equation}
    n.ball.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} R
\end{equation}

\begin{equation}
    radius = |\frac{\sum_{i=1}^{n}d_i}{n} - f(\mathbf{v}d)|
\end{equation}

\begin{equation}
    ranges = |max(d_i) - min(d_i)|
\end{equation}

\begin{equation}
    variances = \sigma^{2}{d_i}
\end{equation}

\begin{equation}
    span.tree.length = \mathrm{branch\ length}
\end{equation}

Where *d* is the number of dimensions,
*n* the number of elements,
$\Gamma$ is the Gamma distribution,
$\lambda_i$ is the eigenvalue of each dimensions,
$\sigma^{2}$ is their variance and
$Centroid_{k}$ is their mean,
$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid),
and $Reference_{k}$ is an arbitrary point's coordinates (usually 0).

@@

## Calculating disparity as the sum and
## the product of these variances
summary(dispRity(dummy_space, metric = c(sum, variances)))
summary(dispRity(dummy_space, metric = c(prod, variances)))

## Calculating the quantiles of each dimension
## in the ordinated space
quantiles(dummy_space)

## Calculating disparity as the distribution of these variances
summary(dispRity(dummy_space, metric = quantiles))

## By default, the quantile calculated is the 95%
## (i.e. 95% of the data on each axis)
## this can be changed using the option quantile:
summary(dispRity(dummy_space, metric = quantiles, quantile = 50))

@@

#### Centroids, displacements, and ancestral distance metrics {#centroids}

@@

```{r}
## The distance between each element and the origin
## of the ordinated space
summary(dispRity(dummy_space, metric = centroids, centroid = 0))

## Disparity as the distance between each element
## and a specific point in space
summary(dispRity(dummy_space, metric = centroids,
                 centroid = c(0,1,2,3,4)))
```

If you have subsets in your `dispRity` object, you can also use the `matrix.dispRity` (see [utilities](#utilities)) and `colMeans` to get the centre of a specific subgroup.
For example

```{r, warning = FALSE}
## Create a custom subsets object
dummy_groups <- custom.subsets(dummy_space,
                               group = list("group1" = 1:5,
                                            "group2" = 6:10))
summary(dispRity(dummy_groups, metric = centroids,
    centroid = colMeans(matrix.dispRity(dummy_groups, "group1"))))
```

The `displacements` distance is the ratio between the `centroids` distance and the `centroids` distance with `centroid = 0`.
Note that it is possible to measure a ratio from another point than `0` using the `reference` argument.
It gives indication of the relative displacement of elements in the multidimensional space: a score  >1 signifies a displacement *away* from the reference. A score of >1 signifies a displacement *towards* the reference.

```{r}
## The relative displacement of the group in space to the centre
summary(dispRity(dummy_space, metric = displacements))

## The relative displacement of the group to an arbitrary point
summary(dispRity(dummy_space, metric = displacements,
                 reference = c(0,1,2,3,4)))
```

@@

#### Functional divergence and evenness

The `func.div` and `func.eve` functions are based on the `FD::dpFD` package.
They are the equivalent to `FD::dpFD(matrix)$FDiv` and `FD::dpFD(matrix)$FEve` but a bit faster (since they don't deal with abundance data).
They are pretty straightforward to use:

```{r}
## The ratio of deviation from the centroid 
summary(dispRity(dummy_space, metric = func.div))

## The minimal spanning tree distances evenness
summary(dispRity(dummy_space, metric = func.eve))

## The minimal spanning tree manhanttan distances evenness
summary(dispRity(dummy_space, metric = func.eve,
                 method = "manhattan"))
```

#### Orientation: angles and deviations

The `angles` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension. This slope can be converted into different units, `"slope"`, `"degree"` (the default) and `"radian"`. This can be changed through the `unit` argument. 
By default, the angle is measured from the slope 0 (the horizontal line in a 2D plot) but this can be changed through the `base` argument (using the defined `unit`):

```{r}
## The distribution of each angles in degrees for each
## main axis in the matrix
summary(dispRity(dummy_space, metric = angles))

## The distribution of slopes deviating from the 1:1 slope:
summary(dispRity(dummy_space, metric = angles, unit = "slope",
                 base = 1))
```

The `deviations` function is based on a similar algorithm as above but measures the deviation from the main axis (or hyperplane) of variation.
In other words, it finds the least square line (for a 2D dataset), plane (for a 3D dataset) or hyperplane (for a >3D dataset) and measures the shortest distances between every points and the line/plane/hyperplane.
By default, the hyperplane is fitted using the least square algorithm from `stats::glm`:

```{r}
## The distribution of the deviation of each point
## from the least square hyperplane
summary(dispRity(dummy_space, metric = deviations))
```

It is also possible to specify the hyperplane equation through the `hyperplane` equation. The equation must contain the intercept first and then all the slopes and is interpreted as $intercept + Ax + By + ... + Nd = 0$. For example, a 2 line defined as beta + intercept (e.g. $y = 2x + 1$) should be defined as `hyperplane = c(1, 2, 1)` ($2x - y + 1 = 0$).

```{r}
## The distribution of the deviation of each point
## from a slope (with only the two first dimensions)
summary(dispRity(dummy_space[, c(1:2)], metric = deviations,
                 hyperplane = c(1, 2, -1)))
```

Since both the functions `angles` and `deviations` effectively run a `lm` or `glm` to estimate slopes or hyperplanes, it is possible to use the option `significant = TRUE` to only consider slopes or intercepts that have a slope significantly different than zero using an `aov` with a significant threshold of $p = 0.05$.
Note that depending on your dataset, using and `aov` could be completely inappropriate!
In doubt, it's probably better to enter your `base` (for `angles`) or your `hyperplane` (for `deviations`) manually so you're sure you know what the function is measuring.

### Which disparity metric to choose?

The disparity metric that gives the most consistent results is the following one:

```{r}
best.metric <- function() return(42)
```
Joke aside, this is a legitimate question that has no simple answer: **it depends on the dataset and question at hand**.
Thoughts on which metric to choose can be find in @moms and @Guillerme2020 but again, will ultimately depend on the question and dataset.
The question should help figuring out which type of metric is desired: for example, in the question "does the extinction released niches for mammals to evolve", the metric in interest should probably pick up a change in size in the trait space (the release could result in some expansion of the mammalian morphospace); or if the question is "does group X compete with group Y", maybe the metric of interested should pick up changes in position (group X can be displaced by group Y).

In order to visualise what signal different disparity metrics are picking, you can use the [`moms`](https://tguillerme.shinyapps.io/moms/) that come with a detailed [manual](https://rawcdn.githack.com/TGuillerme/moms/master/inst/moms_vignette.html) on how to use it.

Alternatively, you can use the `test.metric` function:

#### `test.metric` {#test-metric}

This function allows to test whether a metric picks different changes in disparity. It intakes the space on which to test the metric, the disparity metric and the type of changes to apply gradually to the space.
Basically this is a type of biased data rarefaction (or non-biased for `"random"`) to see how the metric reacts to specific changes in trait space.

```{r}
# Creating a 2D uniform space
example_space <- space.maker(300, 2, runif)

## Testing the product of ranges metric on the example space
example_test <- test.metric(example_space, metric = c(prod, ranges),
                           shifts = c("random", "size")) 
```

By default, the test runs three replicates of space reduction as described in @moms by gradually removing 10% of the data points following the different algorithms from @moms (here the `"random"` reduction and the `"size"`) reduction, resulting in a `dispRity` object that can be summarised or plotted.
The number of replicates can be changed using the `replicates` option.
Still by default, the function then runs a linear model on the simulated data to measure some potential trend in the changes in disparity. 
The model can be changed using the `model` option.
Finally, the function runs 10 reductions by default from keeping 10% of the data (removing 90%) and way up to keeping 100% of the data (removing 0%).
This can be changed using the `steps` option.
A good disparity metric for your dataset will typically have no trend in the `"random"` reduction (the metric is ideally not affected by sample size) but should have a trend for the reduction of interest.

```{r}
## The results as a dispRity object
example_test

## Summarising these results
summary(example_test)

## Or visualising them
plot(example_test)
```

@@

### `type = preview`

Note that all the options above are plotting disparity objects for which a disparity metric *has been calculated*.
This makes totally sense for `dispRity` objects but sometimes it might be interesting to look at what the trait-space looks like before measuring the disparity.
This can be done by plotting `dispRity` objects with no calculated disparity!

For example, we might be interested in looking at how the distribution of elements change as a function of the distributions of different sub-settings.
For example custom subsets *vs.* time subsets:

```{r, fig.width=8, fig.height=4}
## Making the different subsets
cust_subsets <- custom.subsets(BeckLee_mat99,
                               crown.stem(BeckLee_tree,
                                          inc.nodes = TRUE))
time_subsets <- chrono.subsets(BeckLee_mat99,
                               tree = BeckLee_tree,
                               method = "discrete",
                               time = 5)

## Note that no disparity has been calculated here:
is.null(cust_subsets$disparity)
is.null(time_subsets$disparity)

## But we can still plot both spaces by using the default plot functions
par(mfrow = c(1,2))
## Default plotting
plot(cust_subsets)
## Plotting with more arguments
plot(time_subsets, dimensions = c(4,5),
     main = "Some \"low\" dimensions")
```

> DISCLAIMER: This functionality can be handy for exploring the data (e.g. to visually check whether the subset attribution worked) but it might be misleading on how the data is *actually* distributed in the multidimensional space!
Groups that don't overlap on two set dimensions can totally overlap in all other dimensions!

For `dispRity` objects that do contain disparity data, the default option is to plot your disparity data.
However you can always force the `preview` option using the following:

```{r}
par(mfrow = c(2,1))
## Default plotting
plot(disparity_time_slices, main = "Disparity through time")
## Plotting with more arguments
plot(disparity_time_slices, type = "preview",
     main = "Two first dimensions of the trait space")
```

@@

## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Plotting the results
plot(test_unif, main = "Uniform (0,1)")
plot(test_norm1, main = "Normal (0,1)")
plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
plot(test_norm3, main = "Normal (variable + correlation)")
```

@@

---
title: "other functionalities"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: [../References.bib, ../packages.bib]
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{other functionalities}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Other functionalities

The `dispRity` package also contains several other functions that are not specific to multidimensional analysis but that are often used by `dispRity` internal functions.
However, we decided to make these functions also available at a user level since they can be handy for certain specific operations!
You'll find a brief description of each of them (alphabetically) here:

## `char.diff`

This is yet another function for calculating distance matrices.
There are many functions for calculating pairwise distance matrices in `R` (`stats::dist`, `vegan::vegdist`, `cluster::daisy` or `Claddis::MorphDistMatrix`) but this one is the `dispRity` one.
It is slightly different to the ones mentioned above (though not that dissimilar from `Claddis::MorphDistMatrix`) in the fact that it focuses on comparing discrete morphological characters and tries to solve all the problems linked to these kind of matrices (especially dealing with special tokens).

The function intakes a matrix with either `numeric` or `integer` (`NA` included) or matrices with `character` that are indeed `integers` (e.g.`"0"` and `"1"`).
It then uses a [bitwise operations](https://en.wikipedia.org/wiki/Bitwise_operations_in_C) architecture implemented in `C` that renders the function pretty fast and pretty modular.
This bitwise operations translates the character states into binary values.
This way, `0` becomes `1`, `1` becomes `2`, `2` becomes `4`, `3` becomes `8`, etc...
Specifically it can handle any rules specific to special tokens (i.e. symbols) for discrete morphological characters. For example, should you treat missing values `"?"` as `NA` (ignoring them) or as any possible character state (e.g. `c("0", "1")`?)? And how to treat characters with a ampersand (`"&"`)? `char.diff` can answer to all these questions!

Let's start by a basic binary matrix 4\*3 with random `integer`:

```{r}
## A random binary matrix
matrix_binary <- matrix(sample(c(0,1), 12, replace = TRUE), ncol = 4,
                        dimnames = list(letters[1:3], LETTERS[1:4]))
```

By default, `char.diff` measures the hamming distance between characters:

```{r}
## The hamming distance between characters
(differences <- char.diff(matrix_binary))
```

Note that the results is just a pairwise distance (dissimilarity) matrix with some special dual class `matrix` and `char.diff`.
This means it can easily be plotted via the disparity package:

```{r}
## Visualising the matrix
plot(differences)
```

You can check all the numerous plotting options in the `?plot.char.diff` manual (it won't be developed here).

The `char.diff` function has much more options however (see all of them in the `?char.diff` manual) for example to measure different differences (via `method`) or making the comparison work per row (for a distance matrix between the rows):

```{r}
## Euclidean distance between rows
char.diff(matrix_binary, by.col = FALSE, method = "euclidean")
```

We can however make it more interesting by playing with the different rules to play with different tokens.
First let's create a matrix with morphological characters as numeric `character`s:

```{r}
## A random character matrix
(matrix_character <- matrix(sample(c("0","1","2"), 30, replace = TRUE), ncol = 5,
                           dimnames = list(letters[1:6], LETTERS[1:5])))

## The hamming difference between columns
char.diff(matrix_character)
```

Here the characters are automatically converted into bitwise integers to be compared efficiently. We can now add some more special tokens like `"?"` or `"0/1"` for uncertainties between state `"0"` and `"1"` but not `"2"`:

```{r}
## Adding uncertain characters
matrix_character[sample(1:30, 8)] <- "0/1"

## Adding missing data
(matrix_character[sample(1:30, 5)] <- "?")

## The hamming difference between columns including the special characters
char.diff(matrix_character)
```

Note here that it detected the default behaviours for the special tokens `"?"` and `"/"`: `"?"` are treated as `NA` (not compared) and `"/"` are treated as both states (e.g. `"0/1"` is treated as `"0"` and as `"1"`).
We can specify both the special tokens and the special behaviours to consider via `special.tokens` and `special.behaviours`.
The `special.tokens` are `missing = "?"`, `inapplicable = "-"`, `uncertainty = "\"` and `polymorphism = "&"` meaning we don't have to modify them for now.
However, say we want to change the behaviour for `"?"` and treat them as all possible characters and treat `"/"` as only the character `"0"` (as an integer) we can specify them giving a behaviour function:

```{r}
## Specifying some special behaviours
my_special_behaviours <- list(missing = function(x,y) return(y),
                              uncertainty = function(x,y) return(as.integer(0)))

## Passing these special behaviours to the char.diff function
char.diff(matrix_character, special.behaviour = my_special_behaviours)
```

The results are quiet different as before! Note that you can also specify some really specific behaviours for any type of special token.

```{r}
## Adding weird tokens to the matrix
matrix_character[sample(1:30, 8)] <- "%"

## Specify the new token and the new behaviour
char.diff(matrix_character, special.tokens = c(weird_one = "%"),
                            special.behaviours = list(
                                weird_one = function(x,y) return(as.integer(42)))
                            )
```

Of course the results can be quiet surprising then... But that's the essence of the modularity. You can see more options in the function manual `?char.diff`!

## `clean.data`

This is a rather useful function that allows matching a `matrix` or a `data.frame` to a tree (`phylo`) or a distribution of trees (`multiPhylo`).
This function outputs the cleaned data and trees (if cleaning was needed) and a list of dropped rows and tips.

```{r}
## Generating a trees with labels from a to e
dummy_tree <- rtree(5, tip.label = LETTERS[1:5])

## Generating a matrix with rows from b to f
dummy_data <- matrix(1, 5, 2, dimnames = list(LETTERS[2:6], c("var1", "var2")))

##Cleaning the trees and the data
(cleaned <- clean.data(data = dummy_data, tree = dummy_tree))
```

## `crown.stem`

This function quiet handily separates tips from a phylogeny between crown members (the living taxa and their descendants) and their stem members (the fossil taxa without any living relatives).

```{r}
data(BeckLee_tree)
## Diving both crow and stem species
(crown.stem(BeckLee_tree, inc.nodes = FALSE))
```

Note that it is possible to include or exclude nodes from the output.
To see a more applied example: this function is used in [chapter 03: specific tutorials](#custom-subsets).

## `get.bin.ages`

This function is similar than the `crown.stem` one as it is based on a tree but this one outputs the stratigraphic bins ages that the tree is covering.
This can be useful to generate precise bin ages for the `chrono.subsets` function:

```{r}
get.bin.ages(BeckLee_tree)
```

Note that this function outputs the stratigraphic age limits by default but this can be customisable by specifying the `type` of data (e.g. `type = "Eon"` for eons).
The function also intakes several optional arguments such as whether to output the startm end, range or midpoint of the stratigraphy or the year of reference of the International Commission of Stratigraphy.
To see a more applied example: this function is used in [chapter 03: specific tutorials](#chrono-subsets).

## `pair.plot`

This utility function allows to plot a matrix image of pairwise comparisons.
This can be useful when getting pairwise comparisons and if you'd like to see at a glance which pairs of comparisons have high or low values.

```{r}
## Random data
data <- matrix(data = runif(42), ncol = 2)

## Plotting the first column as a pairwise comparisons
pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = 1)
```

Here blue squares are ones that have a high value and orange ones the ones that have low values.
Note that the values plotted correspond the first column of the data as designated by `what = 1`.

It is also possible to add some tokens or symbols to quickly highlight to specific cells, for example which elements in the data are below a certain value:

```{r}
## The same plot as before without the diagonal being the maximal observed value
pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = "max")
## Highlighting with an asterisk which squares have a value below 0.2
pair.plot(data, what = 1, binary = 0.2, add = "*", cex = 2)
```

This function can also be used as a binary display when running a series of pairwise t-tests.
For example, the following script runs a wilcoxon test between the time-slices from the `disparity` example dataset and displays in black which pairs of slices have a p-value below 0.05:

```{r}
## Loading disparity data
data(disparity)

## Testing the pairwise difference between slices
tests <- test.dispRity(disparity, test = wilcox.test, correction = "bonferroni")

## Plotting the significance
pair.plot(as.data.frame(tests), what = "p.value", binary = 0.05)
```

## `reduce.matrix`

This function allows to reduce columns or rows of a matrix to make sure that there is enough overlap for further analysis.
This is particularly useful if you are going to use distance matrices since it uses the `vegan::vegdist` function to test whether distances can be calculated or not.

For example, if we have a patchy matrix like so (where the black squares represent available data):

```{r}
set.seed(1)
## A 10*5 matrix
na_matrix <- matrix(rnorm(50), 10, 5)
## Making sure some rows don't overlap
na_matrix[1, 1:2] <- NA
na_matrix[2, 3:5] <- NA
## Adding 50% NAs
na_matrix[sample(1:50, 25)] <- NA
## Illustrating the gappy matrix
image(t(na_matrix), col = "black")
```

We can use the `reduce.matrix` to double check whether any rows cannot be compared.
The functions needs as an input the type of distance that will be used, say a `"gower"` distance:

```{r}
## Reducing the matrix by row
(reduction <- reduce.matrix(na_matrix, distance = "gower"))
```

We can not remove the rows 1 and 9 and see if that improved the overlap:

```{r}
image(t(na_matrix[-as.numeric(reduction$rows.to.remove), ]), col = "black")
```

## `slice.tree`

This function is a modification of the `paleotree::timeSliceTree` function that allows to make slices through a phylogenetic tree.
Compared to the `paleotree::timeSliceTree`, this function allows a model to decide which tip or node to use when slicing through a branch (whereas `paleotree::timeSliceTree` always choose the first available tip alphabetically).
The models for choosing which tip or node are the same as the ones used in the `chrono.subsets` and are described in [chapter 03: specific tutorials](#chrono-subsets).

The function works by using at least a tree, a slice age and a model:

```{r}
set.seed(1)
## Generate a random ultrametric tree
tree <- rcoal(20)
## Add some node labels
tree$node.label <- letters[1:19]
## Add its root time
tree$root.time <- max(tree.age(tree)$ages)

## Slicing the tree at age 0.75
tree_75 <- slice.tree(tree, age = 0.75, "acctran")

## Showing both trees
par(mfrow = c(1,2))
plot(tree, main = "original tree")
axisPhylo() ; nodelabels(tree$node.label, cex = 0.8)
abline(v = (max(tree.age(tree)$ages) - 0.75), col = "red")
plot(tree_75, main = "sliced tree")
```

## `slide.nodes` and `remove.zero.brlen`

This function allows to slide nodes along a tree!
In other words it allows to change the branch length leading to a node without modifying the overall tree shape.
This can be useful to add some value to 0 branch lengths for example.

The function works by taking a node (or a list of nodes), a tree and a sliding value.
The node will be moved "up" (towards the tips) for the given sliding value.
You can move the node "down" (towards the roots) using a negative value.

```{r}
set.seed(42)
## Generating simple coalescent tree
tree <- rcoal(5)

## Sliding node 8 up and down
tree_slide_up <- slide.nodes(8, tree, slide = 0.075)
tree_slide_down <- slide.nodes(8, tree, slide = -0.075)

## Display the results
par(mfrow = c(3,1))
plot(tree, main = "original tree") ; axisPhylo() ; nodelabels()
plot(tree_slide_up, main = "slide up!") ; axisPhylo() ; nodelabels()
plot(tree_slide_down, main = "slide down!") ; axisPhylo() ; nodelabels()
```

The `remove.zero.brlen` is a "clever" wrapping function that uses the `slide.nodes` function to stochastically remove zero branch lengths across a whole tree.
This function will slide nodes up or down in successive postorder traversals (i.e. going down the tree clade by clade) in order to minimise the number of nodes to slide while making sure there are no silly negative branch lengths produced!
By default it is trying to slide the nodes using 1% of the minimum branch length to avoid changing the topology too much.

```{r}
set.seed(42)
## Generating a tree
tree <- rtree(20)

## Adding some zero branch lengths (5)
tree$edge.length[sample(1:Nedge(tree), 5)] <- 0

## And now removing these zero branch lengths!
tree_no_zero <- remove.zero.brlen(tree)

## Exaggerating the removal (to make it visible)
tree_exaggerated <- remove.zero.brlen(tree, slide = 1)

## Check the differences
any(tree$edge.length == 0)
any(tree_no_zero$edge.length == 0)
any(tree_exaggerated$edge.length == 0)

## Display the results
par(mfrow = c(3,1))
plot(tree, main = "with zero edges")
plot(tree_no_zero, main = "without zero edges!")
plot(tree_exaggerated, main = "with longer edges")
```

## `tree.age`

This function allows to quickly calculate the ages of each tips and nodes present in a tree.

```{r}
set.seed(1)
tree <- rtree(10)
## The tree age from a 10 tip tree
tree.age(tree)
```

It also allows to set the age of the root of the tree:

```{r}
## The ages starting from -100 units
tree.age(tree, age = 100)
```

Usually tree age is calculated from the present to the past (e.g. in million years ago) but it is possible to reverse it using the `order = present` option:

```{r}
## The ages in terms of tip/node height
tree.age(tree, order = "present")
```

<!--chapter:end:05_other-functionalities.Rmd-->

@@

## The `dispRity` object content {#disprity-object}

The functions above are utilities to easily and safely access different elements in the `dispRity` object.
Alternatively, of course, each elements can be accessed manually.
Here is an explanation on how it works.
The `dispRity` object is a `list` of two to four elements, each of which are detailed below:

 * `$matrix`: an object of class `list` that contains at least one object of class `matrix`: the full multidimensional space.
 * `$call`: an object of class `list` containing information on the `dispRity` object content.
 * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
 * `$disparity`: an object of class `list` containing the disparity values.

The `dispRity` object is loosely based on `C` structure objects.
In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint: at any time, only one matrix (or list of matrices) is present in the `R` environment rather than multiple copies of it for each subset.

### `$matrix`

This is the multidimensional space, stored in the `R` environment as a `list` object containing one or more `matrix` objects.
Each `matrix` requires row names but not column names (optional).
By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
This element of the `dispRity` object is never modified.

### `$call`
